---
title: "hw04 - Tidy Data & Joins"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gapminder)
library(tidyverse)
library(ggplot2)
library(DT)
library(knitr)
```

## Exercise 1
### Univariate Option 1:
1. Make a tibble with one row per year, and columns for life expectancy for two or more countries.

```{r}
#Life expectancy in Iran and Japan for each year in gapminder
gap_wide <- gapminder %>% 
  filter(country == "Iran" | country == "Japan") %>% 
  pivot_wider(id_cols = year, 
              names_from = country, 
              values_from = lifeExp)
DT::datatable(gap_wide)
```


2. Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

```{r}

#Scatterplot of life expectancy in Iran vs Japan

gap_wide %>% 
  ggplot(aes(x = Iran, y = Japan, colour = year)) +
           geom_point() +
           labs(
             title = "Life Expectancy of Iran vs Japan",
             x = "Life Expectancy in Iran",
             y = "Life Expectancy in Japan"
           )

```


3. Re-lengthen the data.


```{r}

#re-lengthen data
gap_wide %>% 
  pivot_longer(cols = c(-year),
               names_to = 'country',
               values_to = 'lifeExp') %>% 
  DT::datatable()

```

## Exercise 2
### Multivariate Option 1:

1. Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.

```{r}

#widen data

newgap_wide <- gapminder %>% 
  filter(country == "Cambodia" | country == "Rwanda") %>% 
  pivot_wider(id_cols     = year, 
              names_from  = country, 
              names_sep   = " ", 
              values_from = c(lifeExp, gdpPercap))

DT::datatable(newgap_wide)

```


2. Re-lengthen the data.

```{r}
#re-lengthen data

newgap_wide %>% 
  pivot_longer(cols = c(-year),
               names_to = c(".value", 'country'),
               names_sep = " ") %>% 
  DT::datatable(newgap_wide)

```

## Exercise 3

```{r}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```


### 3.1
For each guest in the guestlist (`guest` tibble), add a column for email address, which can be found in the `email` tibble.

```{r}

#lengthen listed guest names and change column name to match 'guest' dataframe
email <- email %>%
  rename(name = guest) %>%
  separate_rows(name, sep = ", ")

#join data tables
guest  %>%
  left_join(email, by = "name") %>%

DT::datatable(guest)
  
```

### 3.2
Who do we have emails for, yet are not on the guestlist?

```{r}
#datatable with values excluded in 'guest'
email %>% 
  anti_join(guest, by = "name") %>% 
  knitr::kable()
```


### 3.3
Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).

```{r}
#include all individuals from the email datatable in the guest datatable
guest %>% 
  full_join(email) %>% 
  DT::datatable()
```

